The markings of the trade: exploring the patterns of olive oil production in Roman Baetica

### 2020-08-21 Paper/jaccardbaetica.tex
correction until measuring the dissimilarity

\xavi{me parece que aquí falta más chicha, porque saltas de una cosa super general (complejidad, economía, grandes dinámicas) a unos casos muy específicos. Deberías aquí explicar algo más sobre tu estudio y lo que quieres entender, por ejemplo qué preguntas específicas quieres responder? Como se enlazan con lo que has explicado hasta ahora? Cómo es que ahora saltas a casos de estudio regionales? Las preguntas específicas de los 3 casos de estudio deberían entrelazarse en una pregunta general del paper, que esté vinculada a lo que has explicado hasta ahora}

@@

The landscape of the \textit{Baetican} province was one of the best regions to face the increase in demand of olive oil across the Roman provinces. For this reason the area saw an increase in productivity as a massive infrastructure of olive oil production was gradually deployed. The production and distribution of olive oil increased from the first to the third centuries AD ~\citep{remesal_concierto}.

Hundreds of presses and amphora-making workshops were build near large extensions covered by olive trees. The workshops where amphorae were made and filled with olive oil were places along the rivers Guadalquivir and Genil riverbanks (see Fig.\ref{workshop}).

@@

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/baetica}
\caption{Distribution of stamps found in Dressel-20 workshops along the Guadalquivir riverbank. The size of each point depends on the number of stamps found on each site}

@@

%\xavi{deberías desarrollar esto un poco más; qué tipo de evidencia demuestra esto? pocos cambios en qué? los talleres? las tipologías anfóricas?} .

@@

Dressel 20 amphorae have a recognisable globular form with short neck and oval-shaped handles. Its capacity allowed to hold 70-75 litres of olive oil to be shipped and distributed throughout Empire \citep{berni_dressel_2016}.

%A large majority of this Baetican amphorae production is composed by Dressel 20 types\citep{dressel_ricerche_1878,
%millet_anforas_1998}.

%were shipped and distributed from \textit{Baetica} throughout western Europe. 
%\xavi{esto ya lo has mencionado; quizás podrías explicar en 2 líneas la capacidad y forma de la dressel 20}.

@@

Dressel 20 was the amphora most stamped during the Roman Empire \citep[18]{millet_anforas_1998}.

@@

The large sample size of recovered Dressel 20 stamps is distributed across all the western provinces of the Roman Empire. For this reason there are several publications of this type of evidence discussing its origin, its meaning, its long period of use and its spatial distribution~\citep{dressel_ricerche_1878,
rodriguez_economioleicola_1977,
chicepi1985,millet_anforas_1998, remesal_sellar_2016} (Fig.\ref{amphora}).

%\xavi{si dices que es muy popular debería haber aquí más citas aparte de Remesal}.

@@

Most stamp codes display a large activity of production that can be difficult to date using an accurate chronology. They have been frequently dated with the consular dating by studying \textit{tituli picti} found in \textit{Monte Testaccio} \citep{Testaccio1, berni_millet_epigrafianforica_2008}.
Despite \textit{tituli picti} contain important information about chronology, they have been still found in few quantities mostly concentrated in a only settlement such as \textit{Monte Testaccio}. In addition, the chronology of stamps could be also biased taking into account that amphorae were deposited after being produced and consumed\xavi{esto necesita un poco de desarrollo. quieres decir que el código se encuentra en contextos de cronologías diferentes no? Y lo del Testaccio, pos eso, que hay pocos y todos de un yacimioento, lo que añade un bias}. 
\maria{desarrollado}

The use of amphoric stamps can be a good proxy to explore the system of olive oil production and trade developed in \textit{Baetica} province. These stamps are typically found on handles but they were also imprinted on rims and the amphora body~\citep{millet_anforas_1998}. The content information of the stamps is shown in different forms and letter content and it seems that there was not a unique criterion.
%\xavi{qué information? El código de letras?}. 
Stamps often displayed a code of three letters and they are known as \textit{Tria Nomina} \citep{berni_millet_amphora_1996}. The three letters correspond to a initial forms of a name but they can also appear complete. 
%\xavi{que quieres decir con abbreviated o complete? Como abrevias un código de 3 letras?}

@@

The stamp codes are interpreted based on three main ideas: content (olive oil), context (amphora workshop) and subject (individuals involved on the production). On the one hand, it seems that stamps could have been identified as the landowner of the olive groves \citep{rodriguez_economioleicola_1977}. On the other hand, they could also belong to the owner of the workshops were the amphorae was made or even to a group of amphora workers~\citep{berni_millet_epigrafianforica_2008}. 
%In any case, the use of these stamps is a good proxy to explore the system of olive oil production and trade developed in Baetica.

\xavi{acabas de decir que tampoco se sabe qué es el sello, así que esta frase queda rara} 
\maria{he cambiado la frase de sitio para que se contextualice mejor}
Nevertheless, some challenges remain under discussion such as how this production was organised and whether it is possible to distinguish production patterns in the olive oil trade. Our questions will be focused on the distribution of amphoric stamps. Did they follow a distribution pattern? Did stamps share the same workshop?

%Neither the use of written records have allowed providing enough information that can explain the economic role of \textit{Baetica} province in the Roman organisation.\xavi{esta frase esta fuera de contexto. Si la pones entonces deberías expandir la idea de frase a párrafo, y poner referencias}.
\maria{he eliminado la última frase}

@@

The Roman conquest in \textit{Britannia} and \textit{Germania} was a new opening to foreign trade in order to supply the military campaign. A high demand for olive oil in both provinces stimulated an important trade network involving different agents and the Empire itself. \textit{Baetica} became a strategic province for the production and distribution in both provinces since it can be detected by the high presence of amphoric stamps found in civil and military areas during this period.

@@

The evidence for olive oil consumption in \textit{Britannia} is scarce before the Roman conquest \citep{funari_corpus_1996,carreras_abastecimiento_2003}. Local population did not consume this product and it was not produced on the region as the landscape and climate of the British isles was not suitable for olive oil trees\citep[161]{monfort_britanniaen_1998}. Based on the archaeological records, this absence of olive oil changed with the arrival of a large amount of Dressel 20 amphorae to \textit{Britannia} from the first century A.D \citep{peacock_amphorae_1991,
carreras_britannia_1998}. At this moment, we detect an increase of the olive oil exportation concurring with the displacement of legions during the military campaigns \citep[161]{monfort_britanniaen_1998}
%\xavi{en qué siglo?}

The increase of the exportation of Dressel 20 amphorae created an important commercial network for exchanges. The demand of olive oil after the Roman conquest required of a new trade route due to the lack of local olive oil production; \textit{Baetica} was chosen as its main supply.

@@

The presence of Dressel 20 stamps in military camps in \textit{Britannia} has been widely studied in Roman archaeology \citep{williams_importation_1983,
funari_corpus_1996,
carreras_britannia_1998,carreras_abastecimiento_2003}.%\xavi{si dices widely studied no puedes referencias solo 1 autor}.

This intense consumption of olive oil also indicates a possible provincial structure designed to organize the supply of olive oil to military camps,

@@

%\xavi{pero entonces la dressel 23 de donde viene? Cuidao que estás diciendo que se acaban las dressel 20, pero esto implica que hay menos aceite de oliva o que viene de otras partes o que va en recipientes distintos a la dressel 20?}

@@

%Julius Caesar's campaigns in \textit{Gallia} during 1st BC opened the door to the invasion of \textit{Germania} during Augustus rule~\citep{remesal_annona_1986,remesal_baetica_2002}\xavi{no hay un resumen de la conquista similar para Britannia, queda raro}. 
\maria{lo he eliminado}

@@

The massive importation of Dressel 20 in both \textit{Germania} provinces could be explained due to the archaeological sources found mostly in German \textit{limes}. 
By contrast, it is still unknown if trade agents participated on the distribution of Baetican olive oil in \textit{Germania} or the supply was exclusively focused on the Roman army garrisons\citep[156]{remesal_germn_2010}
\xavi{quieres decir que aquí sólo encontramos dressel 20 en yacimientos militares porque no se han excavado muchos civiles?}.
\maria{lo he mejorado a ver si se entiende}

@@

There databased allowed us to retrieve 3798 stamps found in Dressel 20 amphorae recovered from workshops within \textit{Baetica} province. Each of these stamps ha detailed records on the site where it was found, the inscribed code as well as its spatial coordinates. The stamps with incomplete information were discarded, thus finishing with a sample of 987 stamps from 81 sites and displaying 130 different code.

@@

According to previous studies those workshops became the most important centres of amphora production in the region \citep{rodriguez_economioleicola_1977,
arva_1997}.
%\xavi{cita de estos estudios}

@@

%\xavi{pero esto no se hizo en el de Baetica no? Por qué?}.

@@

%We analysed a dataset of 2219 stamps from different centres in \textit{Britannia}. 
%CITAR (Callender, 1965; Carreras Monfort y Funari,1998; Ayllón-Martı́ et al., 2018). 
In \textit{Britannia}, we studied a total sample of 1765 stamps from 46 sites displaying 968 unique codes

%\xavi{pero esto no concuerda con las 2219 de antes...es por descartar los inciertos? Hay que dejarlo claro}.

@@

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/britannia}
\caption{Sites in \textit{Britannia} where Dressel-20 amphoric stamps have been found. Each dot is a site while its size shows the sample size of stamps found in the site. A majority of stamps have been found in garrisons related to Hadrian's and Antonine walls}
\label{britannia}
\end{figure}

@@

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/germania}
\caption{Sites in Germania where Dressel-20 amphoric stamps have been found. Each dot is a site while its size shows the sample size of stamps found in the site. Most sites are located along the German \textit{limes}}
\label{germania}
\end{figure}

>>>>

\xavi{me parece que aquí falta más chicha, porque saltas de una cosa super general (complejidad, economía, grandes dinámicas) a unos casos muy específicos. Deberías aquí explicar algo más sobre tu estudio y lo que quieres entender, por ejemplo qué preguntas específicas quieres responder? Como se enlazan con lo que has explicado hasta ahora? Cómo es que ahora saltas a casos de estudio regionales? Las preguntas específicas de los 3 casos de estudio deberían entrelazarse en una pregunta general del paper, que esté vinculada a lo que has explicado hasta ahora}
\maria{he añadido unos párrafos :-)}

@@

The landscape of the \textit{Baetican} province was one of the best regions to face the increased demand for olive oil across the Roman provinces. For this reason, the area saw an increase in productivity as a massive infrastructure of olive oil production was gradually deployed. The production and distribution of olive oil increased from the first to the third centuries AD ~\citep{remesal_concierto}.

Hundreds of presses and amphora-making workshops were built near large extensions covered by olive trees. The workshops where amphorae were made and filled with olive oil were places along the rivers Guadalquivir and Genil riverbanks (see Fig.\ref{workshop}).

@@

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/baetica}
\caption{Distribution of stamps found in Dressel 20 workshops along the Guadalquivir riverbank. The size of each point depends on the number of stamps found on each site}

@@

\xavi{deberías desarrollar esto un poco más; qué tipo de evidencia demuestra esto? pocos cambios en qué? los talleres? las tipologías anfóricas?}
\maria{hecho}

@@

They have a recognisable globular form with a short neck and oval-shaped handles. Its capacity allowed to hold 70-75 litres of olive oil to be shipped and distributed throughout Empire \citep{berni_dressel_2016}.

\xavi{esto ya lo has mencionado; quizás podrías explicar en 2 líneas la capacidad y forma de la dressel 20}.
\maria{hecho y borrado párrafo}

@@

Dressel 20 was the amphora most stamped during the Roman Empire ~\citep[18]{millet_anforas_1998}.

@@

The large sample size of recovered Dressel 20 stamps is distributed across all the western provinces of the Roman Empire. For this reason, there are several publications of this type of evidence discussing its origin, its meaning, its long period of use and its spatial distribution~\citep{dressel_ricerche_1878,
rodriguez_economioleicola_1977,
chicepi1985,millet_anforas_1998, remesal_sellar_2016} (Fig.\ref{amphora}).

\xavi{si dices que es muy popular debería haber aquí más citas aparte de Remesal}
\maria{he incluido nuevas citas}

@@

Most stamps display a large activity of production that can be difficult to date using an accurate chronology. They have been frequently dated with the consular dating by studying \textit{tituli picti} found in \textit{Monte Testaccio} \citep{Testaccio1, berni_millet_epigrafianforica_2008}.
Despite \textit{tituli picti} contain important information about chronology, they have been still found in few quantities mostly concentrated in an only settlement such as \textit{Monte Testaccio}. In addition, the chronology of stamps could be also biased taking into account that amphorae were deposited after being produced and consumed\xavi{esto necesita un poco de desarrollo. quieres decir que el código se encuentra en contextos de cronologías diferentes no? Y lo del Testaccio, pos eso, que hay pocos y todos de un yacimioento, lo que añade un bias}. 
\maria{desarrollado}

The use of amphoric stamps can be a good proxy to explore the system of olive oil production and trade developed in \textit{Baetica} province. These stamps are typically found on handles but they were also imprinted on rims and the amphora body~\citep{millet_anforas_1998}. The content information of the stamps is shown in different forms and letter content and it seems that there was not a unique criterion.
%\xavi{qué information? El código de letras?}. 
Stamps often displayed a code of three letters and they are known as \textit{Tria Nomina} \citep{berni_millet_amphora_1996}. The three letters correspond to initial forms of a name but they can also appear complete. 
\xavi{que quieres decir con abbreviated o complete? Como abrevias un código de 3 letras?}
\maria{corregido}

@@

The stamp codes are interpreted based on three main ideas: content (olive oil), context (amphora workshop) and subject (individuals involved in the production). On the one hand, it seems that stamps could have been identified as the landowner of the olive groves \citep{rodriguez_economioleicola_1977}. On the other hand, they could also belong to the owner of the workshops were the amphorae were made or even to a group of amphora workers~\citep{berni_millet_epigrafianforica_2008}. 
%In any case, the use of these stamps is a good proxy to explore the system of olive oil production and trade developed in Baetica.

\xavi{acabas de decir que tampoco se sabe qué es el sello, así que esta frase queda rara} 
\maria{he cambiado la frase de sitio para que se contextualice mejor}

Nevertheless, some challenges remain under discussion such as how this production was organised and whether it is possible to distinguish production patterns in the olive oil trade. Our questions will be focused on the distribution of amphoric stamps. Did they follow a distribution pattern? Did stamps share the same workshop?

%Neither the use of written records have allowed providing enough information that can explain the economic role of \textit{Baetica} province in the Roman organisation.\xavi{esta frase esta fuera de contexto. Si la pones entonces deberías expandir la idea de frase a párrafo, y poner referencias}.
%\maria{he eliminado la última frase}

@@

The Roman conquest in \textit{Britannia} and \textit{Germania} was a new opening to foreign trade in order to supply the military campaign. High demand for olive oil in both provinces stimulated an important trade network involving different agents and the Empire itself. \textit{Baetica} became a strategic province for the production and distribution in both provinces since it can be detected by the high presence of amphoric stamps found in civil and military areas during this period.

@@

The evidence for olive oil consumption in \textit{Britannia} is scarce before the Roman conquest \citep{funari_corpus_1996,carreras_abastecimiento_2003}. The local population did not consume this product and it was not produced on the region as the landscape and climate of the British Isles was not suitable for olive oil trees\citep[161]{monfort_britanniaen_1998}. Based on the archaeological records, this absence of olive oil changed with the arrival of a large amount of Dressel 20 amphorae to \textit{Britannia} from the first century A.D \citep{peacock_amphorae_1991,
carreras_britannia_1998}. At this moment, we detect an increase of the olive oil exportation concurring with the displacement of legions during the military campaigns \citep[161]{monfort_britanniaen_1998}
%\xavi{en qué siglo?}

The increase of the exportation of Dressel 20 amphorae created an important commercial network for exchanges. The demand for olive oil after the Roman conquest required of a new trade route due to the lack of local olive oil production; \textit{Baetica} was chosen as its main supply.

@@

The presence of Dressel 20 stamps in military camps in \textit{Britannia} has been widely studied in Roman archaeology \citep{williams_importation_1983,
funari_corpus_1996,
carreras_britannia_1998,carreras_abastecimiento_2003}.\xavi{si dices widely studied no puedes referencias solo 1 autor}
\maria{he añadido más autores}

This intense consumption of olive oil also indicates a possible provincial structure designed to organize the supply of olive oil to military camps.

@@

\xavi{pero entonces la dressel 23 de donde viene? Cuidao que estás diciendo que se acaban las dressel 20, pero esto implica que hay menos aceite de oliva o que viene de otras partes o que va en recipientes distintos a la dressel 20?}
\maria{he eliminado esta frase para que quede más claro}

@@

%Julius Caesar's campaigns in \textit{Gallia} during 1st BC opened the door to the invasion of \textit{Germania} during Augustus rule~\citep{remesal_annona_1986,remesal_baetica_2002ç
\xavi{no hay un resumen de la conquista similar para Britannia, queda raro}. 
\maria{lo he eliminado}

@@

The massive importation of Dressel 20 in both \textit{Germania} provinces could be explained due to the archaeological sources found mostly in German \textit{limes}. 
By contrast, it is still unknown if trade agents participated in the distribution of Baetican olive oil in \textit{Germania} or the supply was exclusively focused on the Roman army garrisons\citep[156]{remesal_germn_2010}
\xavi{quieres decir que aquí sólo encontramos dressel 20 en yacimientos militares porque no se han excavado muchos civiles?}.
\maria{lo he mejorado, a ver si se entiende}

@@

The database allowed us to retrieve 3798 stamps found in Dressel 20 amphorae recovered from workshops within \textit{Baetica} province. Each of these stamps ha detailed records on the site where it was found, the inscribed code as well as its spatial coordinates. The stamps with incomplete information were discarded, thus finishing with a sample of 987 stamps from 81 sites and displaying 130 different code.

@@

According to previous studies, those workshops became the most important centres of amphora production in the region \citep{rodriguez_economioleicola_1977,
arva_1997}.
\xavi{cita de estos estudios}
\maria{citados}

@@

\xavi{pero esto no se hizo en el de Baetica no? Por qué?}
\maria{lo he especificado}

@@

%We analysed a dataset of 2219 stamps from different centres in \textit{Britannia}. 
%CITAR (Callender, 1965; Carreras Monfort y Funari,1998; Ayllón-Martı́ et al., 2018). 
In \textit{Britannia}, we studied a total sample of 1765 stamps from 46 sites displaying 968 unique codes.

\xavi{pero esto no concuerda con las 2219 de antes...es por descartar los inciertos? Hay que dejarlo claro}
\maria{clarificado}

@@

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/britannia}
\caption{Sites in \textit{Britannia} where Dressel 20 amphoric stamps have been found. Each dot is a site while its size shows the sample size of stamps found on the site. A majority of stamps have been found in garrisons related to Hadrian's and Antonine walls}
\label{britannia}
\end{figure}

@@

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/germania}
\caption{Sites in Germania where Dressel-20 amphoric stamps have been found. Each dot is a site while its size shows the sample size of stamps found on the site. Most sites are located along the German \textit{limes}}
\label{germania}
\end{figure}
